---
title: "hmmTMB vignette"
author: "Richard Glennie, Théo Michelot"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{hmmTMB vignette}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE, error = FALSE, warning = FALSE,
  comment = NA
)
```

```{r load-package, echo = FALSE}
library(hmmTMB)
set.seed(342)
```

# Observation distribution

Distributions for the observation process are created using the Dist class. To define a new distribution, we provide its name (as a string), its probability density/mass function (as a function), and the link functions for its parameters (as named lists of functions). For example, for the normal distribution, we have
```{r dist-def}
dist_norm <- Dist$new(
  name = "norm", 
  pdf = dnorm,
  link = list(mean = identity, sd = log),
  invlink = list(mean = identity, sd = exp),
  npar = 2
)
```

The Poisson, gamma, normal, beta, and von Mises distributions are included in the package, and more will be implemented in the future.

The functions n2w and w2n transform parameters of the observation distributions from the natural to the working scale and vice-versa. We can verify that applying the two functions successively returns the original parameter values:
```{r norm-test}
# List of parameters (on the natural scale)
par1 <- list(mean = c(0, 5), sd = c(1, 10))

# Transform to working scale
wpar <- dist_norm$n2w(par1)
wpar

# Transform back to natural scale
par2 <- dist_norm$w2n(wpar)
par2
```

# HMM data

Data objects are defined with the HmmData class. An object can be created from a data frame which includes one column for each observed variable and for each covariate. There are special column names: "ID" is used to identify the time series if several are provided, and "time" gives the times of observation. If the times of observations are provided, gaps are automatically filled with NAs to create a obtain a regular time grid.

```{r create-data}
# Create a dummy data set, with two time series
times1 <- seq(as.POSIXct("2020/01/01 00:00:00"), by = "hour", length = 10)
times2 <- seq(as.POSIXct("2020/02/01 00:00:00"), by = "hour", length = 10)
data <- data.frame(ID = rep(c(1, 2), each = 10),
                   x = rnorm(20), 
                   time = c(times1, times2))

# Remove a few rows to create gaps in the data
data <- data[-c(2:3, 7, 15:18),]

# Define HmmData object
data_prep <- HmmData$new(data = data, interval = "hour")

# The gaps have been filled
data_prep$data()
```

# HMM observation model

The class Observation encapsulates the model for the observed variables.

```{r obs-create}
# Create a dummy data set
times <- seq(as.POSIXct("2020/01/01 00:00:00"), by = "hour", length = 100)
data <- data.frame(ID = rep(1, 100),
                   x = rnorm(100), 
                   time = times)
data_prep <- HmmData$new(data = data)

# List of observation distributions
dists <- list(x = dist_norm)

# List of observation parameters
par <- list(x = list(mean = c(0, 0), sd = c(1, 10)))

# Number of states
n_states <- 2

# Create observation process object
obs_process <- Observation$new(data = data_prep, dists = dists, 
                               n_states = n_states, par = par)
```

The parameters on the working scale are calculated when the object is created:
```{r obs-wpar}
obs_process$wpar()
```

The function plot_dist generates a histogram of the observations with the probability density (or mass) function of the specified distribution.
```{r obs-plot}
# Plot histogram and pdf for variable "x"
obs_process$plot_dist(name = "x")
```

## User-specified observation distribution

_Here we will need to explain how additional distributions can be added by users._

## Future work

 - AIC
 
 - confidence intervals
    + on natural parameters
    + on transformed quantities (e.g. covariate-dependent parameters)
    + run simulation experiments to check coverage
 
 - pseudo-residuals
 
 - stationary state probabilities as functions of covariates
 
 - state probabilities (local decoding)
 
 - additional observation distributions
    + Weibull
    + log-normal
    + wrapped Cauchy
    + Bernoulli...
    
 - zero inflation
 
 - improve parametrisation of mean of circular distributions (avoid numerical issues around -pi and pi)
 
 - allow for input of known states (semi-supervised learning)
 
 - allow for constraints on parameters
    + inequalities, e.g. $\mu_1 < \mu_2$
    + bounds, e.g. $-1 < \mu_1 < 1$
    + fixed value, e.g. $\mu_1 = 0$
    
 - obtain starting values for complex models from simpler nested model (e.g. for smooth effects)
 
 - plotting functions to visualise fitted model
    + time series coloured by states
    + covariate effects...
    
 - update/create documentation for class methods and attributes
 
 - informative error messages in constructors
 
 - deal with NAs in covariates (linear interpolation? nearest value? error message?)
 
 - simulation function, to simulate either from a fitted model or based
 on parameter values given as input. Method of Hmm class?
 
 - stationary = TRUE/FALSE option in MarkovChain class
 