---
title: "hmmTMB user guide"
author: "Théo Michelot, Richard Glennie"
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: yes
  html_document:
    number_sections: yes
bibliography: refs.bib
vignette: |
  %\VignetteIndexEntry{hmmTMB user guide} %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: inline
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
message = FALSE, error = FALSE, warning = FALSE,
comment = NA, cache = TRUE
)
```

```{r set-seed, echo = FALSE}
library(hmmTMB)
set.seed(905)
```

# Introduction

The package hmmTMB implements hidden Markov models (HMMs) with flexible covariate dependence in all model parameters. In this document, we briefly describe the structure of the package, and illustrate its use with several examples. For general background on HMMs, see @zucchini2016 and, for a presentation of HMMs in the context of ecological studies, see @mcclintock2020review.

HMMs are time series models involving two processes: an unobserved state process $(S_t)$ specified as a Markov chain, and an observation process $(Z_t)$. At each time $t = 1, 2, \dots$, the distribution of the observation $Z_t$ depends on the value $S_t$ of the state process. There are therefore two sets of parameters:

  - the state process is formulated in terms of a transition probability matrix (and an initial distribution);
  
  - the observation process is formulated in terms of the state-dependent parameters of the observation distributions. 
  
In hmmTMB, the parameters of the state process and of the observation process can depend on covariates, including linear fixed effects, smooth covariate effects using splines, and i.i.d. normal random effects.

# Package structure

The package hmmTMB uses the R6 framework for object-oriented programming (@chang2019), and it is based on three main classes:

  - MarkovChain: model for state process, including formulas for the transition probabilities
  
  - Observation: model for observation process, including state-dependent distributions, and formulas for the state-dependent observation parameters
  
  - HMM: contains a state process model (i.e., a MarkovChain object) and an observation model (i.e., an Observation object).
  
Model specification, model fitting and model visualisation can all be performed by manipulating objects from those three classes. Following the R6 syntax, an object is created with something like
```{r r6-new}
hidden <- MarkovChain$new(n_states = 2, data = data.frame(z = rnorm(100)))
```
This line of code defines a hidden state model with 2 states and no covariate dependence. Once an object is created, various methods (i.e., functions) can be called to manipulate it, using the following syntax:
```{r r6-method}
hidden$formulas()
```

# Example 1: Elk movement analysis

We illustrate the use of the package on a data set of elk movement (from @morales2004), accessible through the moveHMM package. In this example, we showcase the use of smoothing splines to model the relationship between model parameters and covariates.

```{r load-packs}
# Load packages
library(hmmTMB)
library(moveHMM)
```

In analyses of animal movement, the step lengths and turning angles are modelled to capture features of the speed and directionality of the movement. We derive those variables from the location using the function prepData in moveHMM.

```{r elk-prepdata}
tracks <- prepData(elk_data, type = "UTM", coordNames = c("Easting", "Northing"))

head(tracks)
```

The data set now has columns for the coordinates (x and y), the step lengths and turning angles, the ID (track identifier), and a covariate (distance to water). hmmTMB doesn't allow for zero inflation for observation distributions with positive support (e.g. gamma), so we jitter step lengths with value zero.

```{r elk-jitter}
indz <- which(tracks$step == 0)
tracks$step[indz] <- runif(length(indz), 0, 1)
```

The data frame passed to hmmTMB should include a column for ID if necessary, as well as for any response variables and covariates included in the model.

The class Observation encapsulates the observation process model of the hidden Markov model. To create an object of that class, we need to specify distributions for the response variables, as well as initial parameter values. We could also include covariate dependence in the parameters of the observation distributions. We use gamma distributions for the step lengths, and von Mises distributions for the turning angles.

```{r elk-obs-new}
# Number of states of the HMM 
n_states <- 2

# Observation distributions: gamma for the step lengths, and
# von Mises for the turning angles
dists <- list(step = dist_gamma, angle = dist_vm)

# Initial parameters
mean0 <- c(300, 3000)
sd0 <- c(400, 4000)
par0_obs <- list(step = list(shape = mean0^2/sd0^2,
                             scale = sd0^2/mean0),
                 angle = list(mu = c(3, 0),
                              kappa = c(0.2, 1)))

# Create Observation object
obs <- Observation$new(data = tracks, dists = dists, 
                       n_states = n_states, par = par0_obs)
```

Similarly, we need to create an object of the class MarkovChain for the hidden process model, encapsulating the formulas and parameter values. Here, we formulate the transition probabilities as functions of the distance to water covariate, with smoothing splines specified using the syntax from the package mgcv (@wood2017).

```{r elk-hid-new}
# Formula for transition probabilities
formula <- ~ s(dist_water, k = 5, bs = "ts")

# Create MarkovChain object
hid <- MarkovChain$new(n_states = n_states, 
                       structure = formula,
                       data = tracks)
```

Combining the MarkovChain and Observation object, we create an object of the HMM class.

```{r elk-hmm-new}
hmm <- HMM$new(obs = obs, hidden = hid)
```

We fit the model with the method `fit`, and print the output of the optimiser with `out`. It contains the parameters estimates (fixed effects), the value of the objective function at the optimum, and diagnostics from the optimiser.

```{r elk-fit}
hmm$fit(silent = TRUE)

hmm$out()
```

We can get estimates of the HMM parameters using the `par` method. By default, when there are covariates, it only returns the value of the parameters for the first observation of the data.

```{r elk-par}
hmm$par()
```

The method `predict` can be used to obtain confidence intervals of the model parameters. For example, say that we want 95% confidence intervals for the parameters of the observation process,
```{r elk-ci}
hmm$predict(name = "obspar", level = 0.95)
```

The method viterbi implements the Viterbi algorithm, to estimate the most likely state sequence.
```{r elk-viterbi}
states <- hmm$viterbi()

head(states)
```

Various plotting functions are provided to visualise a fitted model. Using the method `plot_dist`, we can plot histograms of the observed data, overlaid with the estimated state-dependent probability density functions.

```{r elk-plot-dist, fig.width = 5, fig.height = 4, fig.show="hold", out.width="50%"}
# Proportion of states in Viterbi sequence, to weigh pdfs
w <- table(states)/length(states)
hmm$plot_dist("step")
hmm$plot_dist("angle")
```

We can plot the data coloured by the Viterbi-decoded states. All plotting functions return ggplot objects, and can easily be edited using ggplot commands, e.g.\ to change axis labels or colours.

```{r elk-plot-ts, fig.width = 6, fig.height = 3, out.width="70%", fig.align="center"}
hmm$plot_ts("step")
```

```{r els-plot-ts2, fig.width = 6, fig.height = 3, out.width = "70%", fig.align="center"}
hmm$plot_ts("x", "y") + coord_equal() + geom_point(size = 0.5)
```

The function `plot` creates graphs of the HMM parameters as functions of a covariate, here distance to water. For example, to obtain a plot of the transition probabilities:

```{r elk-plot-tpm, fig.width = 6, fig.height = 4, out.width="70%", fig.align="center"}
hmm$plot(name = "tpm", var = "dist_water") + xlab("Distance to water (m)")
```

With the same function, we can also visualise the stationary state probabilities as functions of the distance to water, using the option `name = "delta"`:

```{r elk-plot-stat, fig.width = 7, fig.height = 4, out.width="90%", fig.align="center"}
hmm$plot(name = "delta", var = "dist_water") + 
  xlab("Distance to water (m)")
```

We can simulate from the fitted model with the method `simulate`, for example to check how well the model captures features of the real data. Here, we simulate step lengths and turning angles from the model.

```{r elk-sim, results = 'hide'}
# Number of simulated realisations
n_sim <- 500
# Data frame of covariate values for simulation
new_covs <- data.frame(dist_water = sample(tracks$dist_water, 
                                           replace = TRUE, 
                                           size = n_sim))

# Simulate from fitted model
sim_data <- hmm$simulate(n = n_sim, data = new_covs)
```

The output is a data frame with the same columns as the input, plus columns for the simulated variables, and a column for the simulated states.

```{r elk-sim-2}
head(sim_data)
```

We then derive the corresponding locations to visualise the simulated movement trajectory, and extract the states used in the simulation (stored as an attribute of the output of the `simulate` method).

```{r elk-sim-3, results = 'hide', fig.width = 6, fig.height = 4, out.width="70%", fig.align="center"}
# Get locations from simulated step lengths and turning angles
bearings <- cumsum(sim_data$angle)
sim_dxy <- sim_data$step * cbind(cos(bearings), sin(bearings))
sim_data$x <- cumsum(c(0, sim_dxy[-n_sim, 1]))
sim_data$y <- cumsum(c(0, sim_dxy[-n_sim, 2]))

# Add state to simulated data frame
sim_data$state <- attr(sim_data, "state")

# Plot simulated trajectory
ggplot(sim_data, aes(x, y, col = factor(state), group = NA)) + 
  scale_color_manual("State", values = hmmTMB_cols) +
  geom_point(size = 0.7) + 
  geom_path() +
  coord_equal()
```

The HMM class has a dedicated method for simulation-based model checking, called `gof` (for "goodness-of-fit"). It takes a function as an argument, which returns some statistic of interest from the data set (e.g., some quantile of the observed variable). The method evaluates that function for the observed data set and for a large number of data sets simulated from the fitted model. If the observed value of the statistic is an unusual value for the distribution of simulated statistics, it suggests that that feature of the data was not captured well by the model. Here, we apply the `gof` method for quantiles of the step length distribution:

``` {r elk-gof, fig.width = 9, fig.height = 3, out.width="100%"}
# Function that takes 'data' as input and return the statistic(s) of interest
gof_fn <- function(data) {
    quantile(data$step, c(0.1, 0.5, 0.9), na.rm = TRUE)
}

# Run test
gof_out <- hmm$gof(gof_fn, silent = TRUE)
```

# Example 2: wild haggis movement

We further illustrate the functionalities of hmmTMB with the analysis of the wild haggis tracking data set from @michelot2016. The objective of the study is to investigate the effect of terrain slope on the behaviour of wild haggises, accounting for inter-individual differences.

We load the data from @michelot2016, which contains locations for 15 haggis tracks, as well as slope and temperature measurements,
```{r haggis-read}
URL <- paste0("https://besjournals.onlinelibrary.wiley.com/action/downloadSupplement?",
              "doi=10.1111%2F2041-210X.13066&file=mee313066-sup-0001-rawhaggises.csv")
raw <- read.csv(url(URL))

head(raw)
```

We use the prepData function from moveHMM to calculate step lengths, and we transform ID into a factor variable,
```{r haggis-prep}
data <- prepData(raw, type = "UTM")
data$ID <- factor(data$ID)

head(data)
```

We define the hidden state process as a 2-state Markov chain, with a smooth effect of "slope" (using thin-plate regression splines), and a normal random intercept for "ID" to allow for differences between haggises. We don't include the effect of temperature because @michelot2016 found no clear effect.
```{r haggis-hid}
n_states <- 2
hid <- MarkovChain$new(n_states = n_states,
                       structure = ~s(ID, bs = "re") + s(slope, k = 5, bs = "ts"),
                       data = data)
```

To define the observation model, we specify that the step lengths should be modelled with gamma distributions. We choose initial parameter values for the mean and standard deviation based on data visualisations, and derive initial values for the shape and scale based on that,
```{r haggis-obs}
# Observation distributions
dists <- list(step = "gamma")

# Initial parameter values
mean0 <- c(1, 5)
sd0 <- c(0.5, 3)
par0 <- list(step = list(shape = mean0^2/sd0^2,
                         scale = sd0/mean0^2))

# Define observation model
obs <- Observation$new(data = data, 
                       n_states = n_states,
                       dists = dists, 
                       par = par0)
```

We create the HMM object from the MarkovChain and Observation components, fit it using the fit method, and compute the most likely state sequence using the viterbi method,
```{r haggis-fit}
hmm <- HMM$new(hidden = hid, obs = obs)
hmm$fit(silent = TRUE)
data$state <- hmm$viterbi()
```

We can plot the time series of step lengths, or the tracks, coloured by the estimated state sequence,
```{r haggis-plot_ts, fig.width = 5, fig.height = 4, out.width="50%", fig.align="center"}
hmm$plot_ts("step")
```

```{r haggis-plot-ts2, , fig.width = 6, fig.height = 4, out.width="70%", fig.align="center"}
hmm$plot_ts("x", "y") + 
    geom_point(size = 0.5) + 
    coord_equal()
```

These plots suggest that state 1 captured fast movement behaviour (long step lengths) and state 2 captured slow movement (short step lengths). This can also be seen from the estimated state-dependent distributions of step lengths,
```{r haggis-plot_dist, fig.width = 7, fig.height = 4, out.width="90%", fig.align="center"}
hmm$plot_dist("step")
```

To visualise the effect of slope on the behaviour of wild haggises, we plot the stationary state probabilities as functions of the slope,
```{r haggis-plot_stat, fig.width = 7, fig.height = 4, out.width="90%", fig.align="center"}
hmm$plot(name = "delta", var = "slope")
```

When there are several covariates, the plotting functions fix the other covariates to their mean value (for continuous covariates) or to their first level (for factor covariates), as shown at the top of the plot. This can be changed by specifying the argument `covs`, a data frame with a single row and one named column for each covariate.

We observe the same pattern found by @michelot2016: wild haggises tended to adopt the fast movement state at intermediate slopes (between 5 and 30 degrees), and the slow movement state on very flat or very steep terrains. @michelot2016 compared a linear and a quadratic effect of slope; the smoothing splines used here are more flexible, and the implementation automatically estimates the smoothness of the relationship during model fitting.

A measure of inter-individual heterogeneity is given by the standard deviation of the distribution of random intercepts, and can be obtained using the `vcomp` method,
```{r haggis-vcomp}
hid$vcomp()
```

# References
